# Assignment 6: Analysis of longitudinal data

```{r}
date()
```

### Libraries
```{r}
library(tidyverse)
```

## 1. Analysing RATS data

### Reading the data and converting categorical variables to factors
```{r}
setwd("C:/LocalData/salailja/IODS/IODS-project/")
bprs <- read_csv("data/bprs_long.csv")
rats <- read_csv("data/rats_long.csv")
bprs$treatment <- as.factor(bprs$treatment)
bprs$subject <- as.factor(bprs$subject)
bprs$weeks <- as.factor(bprs$weeks)
rats$ID <- as.factor(rats$ID)
rats$Group <- as.factor(rats$Group)
rats$time <- as.factor(rats$time)
```

### Graphing individual trajectories
```{r}
ggplot(rats, aes(x = time, y = weight, group = ID, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none") + 
  scale_y_continuous(limits = c(min(rats$weight), max(rats$weight)))
```
From here it looks like there are considerable differences in overall weight and variance between the three groups, and also within groups to some point. However, average growth of body weight looks approximately same in each group.

Lets add standardized weight variable and do the same plot again
```{r}
rats <- rats %>%
  group_by(time) %>%
  mutate(stdweight = scale(weight)) %>%
  ungroup()

ggplot(rats, aes(x = time, y = stdweight, group = ID, linetype = ID)) +
  geom_line() +
  scale_linetype_manual(values = rep(1:10, times=4)) +
  facet_grid(. ~ Group, labeller = label_both) +
  theme(legend.position = "none")
```
Interpretation stays pretty much similar, but standardized trajectories differ more from each other than the unstandardized ones.

### Graphing summary measures
```{r}
rats_sum <- rats %>%
  group_by(Group, time) %>%
  summarise(mean = mean(weight), se = sd(weight)/sqrt(length(weight))) %>%
  ungroup()

ggplot(rats_sum, aes(x = time, y = mean, group = Group, linetype = Group, shape = Group)) +
  geom_line() +
  scale_linetype_manual(values = c(1,2,3)) +
  geom_point(size=3) +
  scale_shape_manual(values = c(1,2,3)) +
  geom_errorbar(aes(ymin=mean-se, ymax=mean+se, linetype="1"), width=0.3) +
  theme(legend.position = c(0.8,0.8)) +
  scale_y_continuous(name = "mean(weight) +/- se(weight)")
```
Using summary measures gives similar interpretation than plots using individual trajectories.

### Differences between groups
```{r}
ggplot(rats_sum, aes(x = Group, y = mean)) +
  geom_boxplot() +
  stat_summary(fun = "mean", geom = "point", shape=23, size=4, fill = "white") +
  scale_y_continuous(name = "mean(weight)")
```
From this boxplot we can clearly see the large difference between group 1 and other groups, and the smaller difference between groups 2 and 3.

Let's use ANOVA to test these differences (while being little bit dumb since the differences can be seen very clearly from the plot in this situation)
```{r}
fit <- lm(weight ~ Group, data = rats)
anova(fit)
```
There are clear differences between groups as seen in previous plots.

